---
title: "Figure 2: Stress-induced changes in bacterial communities"
author: "Marcela Aragon"
date: "2025-02-03"
output: html_document
---

#General settings 
##Loading libraries
```{r}

#Set working directory to project directory 
setwd("./")
getwd() #ok

#microbiome data
library(phyloseq)
library(microbiome)
library(metagMisc)
library(metagenomeSeq)#CSS normalization
library(vegan)
library(ape)
library(ANCOMBC) # differential abundance with ANCOM

#data handling
library(dplyr)
library(tidyr)
library(tibble)
library(readxl)
library(purrr) #melts df of lists into a single one
library(varhandle)
library(itol.toolkit) #formats into iTOL
library(stringr) #splits text

#stats
library(picante) #faith's Phylogenetic Diversity
library(stats) #stats
library(DHARMa) #check residuals
library(lme4) #run glms
library(fitdistrplus) #checks data distribution
library(emmeans)
library(rempsyc) #tables
library(magrittr) #pipe with emmeans
library(effectsize)
library(scales)#rounds p value
library(umap)
library(EcolUtils) #pairwise Adonis

#ploting
library(ggplot2)
library(viridis)
library(GGally)
library(ggpubr)
library(RColorBrewer) #nice colors for plots


```

## Global settings for plots 

```{r settings for plots}

# Setting up white background for plots
theme_set(theme_bw())

# Set colors for soil_inocula
color_pathway_induced <- c("#989898", "#E69F00", "#009D64") #control, JA and SA
color_stress <- c("#BFBFBF", "#E69F00", "#009D64") #same but for UMAP
color_experiment <- c("#336699", "#996633") #phytohormone-treated and herbivore-infested

# Set default axis looks
axis_looks <- theme(axis.text.x = element_text(colour = "black", size = 9,
                                               face = "bold", angle=0, hjust=0.5))+ #text in X
  theme(axis.text.y = element_text(colour = "black", size = 9, face = "bold"))+ #text in y
  theme(axis.title=element_text(size=11, face = "bold"))+ #all texts
  theme(axis.title.y = element_text(margin = ggplot2::margin(t = 0, r = 10, b = 0, l = 0)))+
  theme(panel.grid.major = element_blank(), #no grid
        panel.grid.minor = element_blank()) #no grid
  #theme(legend.position="none")+ #no legend

#Get names for boxes in facet wrap
Experiment_names <- c(
                    `simulated` = "Phytohormone-treated",
                    `real_insects` = "Herbivore-infested")

```

#(a)UMAP full 16S community
##Data
```{r 16S UMAP; data}

##Loading data
#16S
load(file = "./Data/phyloseq_objects/ps_16S_full.RData")
ps_16S

#16S split by experiment
load(file = "./Data/phyloseq_objects/ps_16S_ByExperiment.RData")
ps_16S_ByExperiment

#let's rename it to ps_16S_split to not re-write the code
ps_16S_split <- ps_16S_ByExperiment

#remove the extra ps
rm(ps_16S_ByExperiment)

```

```{r UMAP; data filtering}

#let's clean our current ps objects by abundance and prevalence to keep only relevant ASVs for analysis

#Add ASV name in taxonomy table
t <- as.data.frame(tax_table(ps_16S))
t$ASV <- rownames(t)
tax_table(ps_16S) <- tax_table(object = as.matrix(t))

#Add it also to the split ones using lapply to apply the function at the same time to the list of ps objects
ps_16S_split <- lapply(ps_16S_split, function(ps){ 
  t <- as.data.frame(tax_table(ps))
  t$ASV <- rownames(t)
  tax_table(ps) <- tax_table(object = as.matrix(t))
  return(ps)
})

#Apply filtering to remove low prevalence (> 10 samples) and abundant (> 0.001% )ASVs
#to calculate and plot beta-diversity in the UMAP space

#---- By experiment ----
asvs_to_keep <- lapply(ps_16S_split, function(ps){
  
  ra_ps <- transform_sample_counts(ps, function(x) x/sum(x))#change it to relative abundance
  ra_f <- filter_taxa(ra_ps, function(x) mean(x) > 0.00001, prune = TRUE)#filter out less than 0.001%
  p_f <- phyloseq_filter_prevalence(ra_f, prev.trh=0.1, abund.trh=NULL)#filter out not present in <10% samples
  
  asvs <- rownames(p_f@otu_table) #get names of ASVs
  print(length(asvs)) #get how many
  return(asvs)
})

#make a copy
Bac_exp_f <- ps_16S_split

#filter out phyloseq object
Bac_exp_f$cond_phytohormones <- subset_taxa(ps_16S_split$cond_phytohormones,
                                            ASV %in% asvs_to_keep$cond_phytohormones)

Bac_exp_f$cond_real_insects <- subset_taxa(ps_16S_split$cond_real_insects,
                                            ASV %in% asvs_to_keep$cond_real_insects)

#final checks
Bac_exp_f #ok

#see how many reads were kept
sum(taxa_sums(Bac_exp_f$cond_phytohormones))/sum(taxa_sums(ps_16S_split$cond_phytohormones)) #88%
sum(taxa_sums(Bac_exp_f$cond_real_insects))/sum(taxa_sums(ps_16S_split$cond_real_insects)) #91%

#check sparsity 
summarize_phyloseq(ps_16S)[[6]] #97%
lapply(ps_16S_split, function(ps) summarize_phyloseq(ps)[[6]])#92
lapply(Bac_exp_f, function(ps) summarize_phyloseq(ps)[[6]])#65-62, zero-inflation improved with filtering. 

#---- Full ----
#Get ASVs to keep
ra_ps <- transform_sample_counts(ps_16S, function(x) x/sum(x))#change it to relative abundance
ra_f <- filter_taxa(ra_ps, function(x) mean(x) >0.00001, prune = TRUE)#filter out less than 0.0001%
p_f <- phyloseq_filter_prevalence(ra_f, prev.trh=0.1, abund.trh=NULL)#filter out not present in <10% samples
asvs <- rownames(p_f@otu_table) #get names of ASVs
length(asvs) #get how many

#filter out phyloseq object
ps_Bac_f <- subset_taxa(ps_16S, ASV %in% asvs)

#checks
#see how many reads were kept
sum(taxa_sums(ps_Bac_f))/sum(taxa_sums(ps_16S)) #87%

#check sparsity 
summarize_phyloseq(ps_16S)[[6]] #97%
summarize_phyloseq(ps_Bac_f)[[6]] #64%, improved.

#clean
rm(asvs_to_keep, ra_ps, ra_f, p_f, asvs,t)

```

Now, let's normalize using cumulative-sum scaling (CSS)
```{r UMAP; CSS normalization}

#for ps split by experiment
Bac_CSS <- lapply(Bac_exp_f, function(ps){phyloseq_transform_css(ps)})
lapply(Bac_CSS, function(ps){head(otu_table(ps))}) #ok

#same for single ps
ps_Bac_css <- phyloseq_transform_css(ps_Bac_f)
head(otu_table(ps_Bac_css))#ok

```

With normalized CSS-transformed ps object now you can calculate distances 
```{r UMAP; distance matrix}

#---- Full ----
#first re-root tree otherwise calculating UniFrac distances is a problem (https://github.com/joey711/phyloseq/issues/936)
ps_Bac_css@phy_tree <- ape::multi2di(ps_Bac_css@phy_tree) 

#Calculate unifrac distances 
uni_full <- phyloseq::distance(ps_Bac_css, "unifrac") 

#Calculate bray-curtis distances [not used]
bray_full <- phyloseq::distance(ps_Bac_css, "bray")


#---- By experiment ----
# first fix tree
Bac_CSS <- lapply(Bac_CSS, function(ps){
  tree <- ps@phy_tree
  tree2 <- ape::multi2di(tree)
  ps@phy_tree <- tree2
  return(ps)
}) #replace tree for the fixed one

#Now, run unifrac distances
uni <- lapply(Bac_CSS, function(ps){ phyloseq::distance(ps, "unifrac")}) #ok

#Bray-Curtis [not used]
bray <- lapply(Bac_CSS, function(ps){ phyloseq::distance(ps, "bray")})

#Phyloseq by Experiment ps will be ploted and analyzed in (b), for now we focus on the full community (ps_Bac_css).
```

##Plot

```{r UMAP; ploting}

#make a list of the 2 distances to use lapply later
dist <- list(bray=bray_full, uni=uni_full) 

#First get a short version of sample data df to be used as metadata for UMAP
df_info_16S <- sample_data(ps_16S) %>% 
               data.frame() %>% 
               dplyr::select(Experiment, plant_pathway_induced)%>%
               mutate(sample_name=rownames(.))%>%
               filter(Experiment %in% c("cond_real_insects", "cond_phytohormones"))

#call function to calculate UMAP coordenates
get_UMAP_coordenates <- function(x){
  
  #dist is a list of distances previously calculated
  set.seed(123)
  umap <- umap(as.matrix(x))
  umap_scores <- as.data.frame(umap$layout)
  colnames(umap_scores) <- c("UMAP1", "UMAP2")
  umap_scores$sample_name <- rownames(umap_scores)
  umap_scores <- umap_scores %>% relocate(sample_name)
  return(umap_scores)
    
}

#get coordinates from dist list
df_UMAP <- lapply(dist, function(x){ get_UMAP_coordenates(x)})

#Now merge it with df 
df_UMAP <- lapply(df_UMAP, function(df){
  x <- left_join(df, df_info_16S, by="sample_name")
  x <- x %>% relocate(sample_name, Experiment, plant_pathway_induced)
})

#call function to plot the UMAP
plot_UMAP <- function(x){
  #x is df 
  #title is the name of the title in between ""
  set.seed(123)
  p <- ggplot(x, aes(x=UMAP1,y=UMAP2, color=plant_pathway_induced)) +
  geom_point(aes(color=plant_pathway_induced, shape=Experiment),
             size=4, alpha=0.85)+ # individual sample values
  #coord_fixed()+ # same axis scaling
  scale_color_manual(values=color_stress)+
  #ggtitle(title)+
  axis_looks+
  theme(legend.position="none")
  
  return(p)
}

#Plot it
u.uni <- plot_UMAP(df_UMAP$uni)

#check
u.uni

```

##Stats
```{r UMAP; stats}

#I need this to test both experiments together 
ps_Bac_css
df_info_16S

#Is there a difference between the bacterial communities among types of induction (Experiment) and stress treatments?
set.seed(234)
perm_full_uni <- adonis2(formula=distance(ps_Bac_css, "unifrac") 
                          ~ Experiment + plant_pathway_induced,
                     data=as(sample_data(ps_Bac_css), "data.frame"),
                     permutations = 999, by="margin") #no, but barely (P=0.071)

#No pairwise post-hoc

```

#(b)UMAP by Experiment
##Data
```{r UMAP split; data}

#data was already obtained in (a), to plot Unifrac distances independently by experiment we need;
Bac_CSS #ps object CSS-transformed and split by Experiment 
bray #Bray-curtis distance matrix calculated by Experiment 
uni #Unifrac distance matrix calculated by Experiment 

#let's put it into a list per experiment so the functions work
dist_phyto <- list(bray=bray$cond_phytohormones, uni=uni$cond_phytohormones)
dist_insects <- list(bray=bray$cond_real_insects, uni=uni$cond_real_insects) 

#let's also get each experiment in a single ps object 
phyto <- Bac_CSS$cond_phytohormones
insects <-Bac_CSS$cond_real_insects
  
```

##Plot
```{r UMAP split; plot Phytohormone-treated}

#---- Phytohormone-treated ----
#We need
phyto #ps object
dist_phyto #list with distance matrices
df_info_16S #metadata df 

#get UMAP coordinates
df_UMAP <- lapply(dist_phyto, function(x){ get_UMAP_coordenates(x)})

#Now merge it with df 
df_UMAP <- lapply(df_UMAP, function(df){
  x <- left_join(df, df_info_16S, by="sample_name")
  x <- x %>% relocate(sample_name, Experiment, plant_pathway_induced)
})

#calculate centroids 
centroids_UMAP <- lapply(df_UMAP, function(df){
  aggregate(cbind(UMAP1, UMAP2) ~ plant_pathway_induced, data=df, FUN=mean)})

#calculate segments
s.uni <- merge(df_UMAP$uni,centroids_UMAP$uni, by.x="plant_pathway_induced", by.y="plant_pathway_induced",
        suffixes=c("","_centroid"))

#call function to plot UMAP with centroids and segments
plot_UMAP <- function(x,y,z,title){
  #x is df 
  #y is the centroid df
  #z is segment df
  #title is the name of the title in between ""
  
  p <- ggplot(x, aes(x=UMAP1,y=UMAP2, color=plant_pathway_induced)) +
  geom_segment(data = z, # spiders
               mapping = aes(x=UMAP1, y=UMAP2, 
                             xend = UMAP1_centroid, yend = UMAP2_centroid),
               linewidth=0.3) + 
  geom_point(size=1.5)+ # individual sample values
  geom_point(data = y, size = 6, alpha=0.9) + # centroids
  #coord_fixed()+ # same axis scaling
  scale_color_manual(values=color_stress)+
    #ggtitle(title)+
  axis_looks+
  theme(legend.position="none")
  return(p)
}

#Plot it
u.uni.phyto <- plot_UMAP(df_UMAP$uni, centroids_UMAP$uni, s.uni, "Unifrac")

#check
u.uni.phyto

```

```{r UMAP split; Plot Herbivore-infested}

#---- Herbivore-infested ----
#We need
insects #ps object
dist_insects #list with distance matrices
df_info_16S #metadata df 

#get UMAP coordinates
df_UMAP <- lapply(dist_insects, function(x){ get_UMAP_coordenates(x)})

#Now merge it with df 
df_UMAP <- lapply(df_UMAP, function(df){
  x <- left_join(df, df_info_16S, by="sample_name")
  x <- x %>% relocate(sample_name, Experiment, plant_pathway_induced)
})

#calculate centroids 
centroids_UMAP <- lapply(df_UMAP, function(df){
  aggregate(cbind(UMAP1, UMAP2) ~ plant_pathway_induced, data=df, FUN=mean)})

#calculate segments
s.uni <- merge(df_UMAP$uni,centroids_UMAP$uni, by.x="plant_pathway_induced", by.y="plant_pathway_induced",
        suffixes=c("","_centroid"))

#Plot it
u.uni.insects <- plot_UMAP(df_UMAP$uni, centroids_UMAP$uni, s.uni, "Unifrac")

#check
u.uni.insects

```

##Stats
```{r UMAP split; stats}

#I need this to test both experiments independently
Bac_CSS

#Is there a difference between bacterial communities among treatments in each experiment?
set.seed(123)

#Unifrac
perm_Byexp_uni <- lapply(Bac_CSS, function(ps){
          set.seed(123)
          perm <- adonis2(formula=phyloseq::distance(ps, "unifrac") 
                          ~ plant_pathway_induced,
                     data=as(sample_data(ps), "data.frame"),
                     permutations = 999, by="margin")
          perm <- perm %>% rownames_to_column("factor")
          return(perm)
          })

#check
perm_Byexp_uni

```

clean up a bit before continuing to (c)
```{r}
#make list of objects to keep
keep_objects <- c("color_experiment", "color_pathway_induced", "color_stress", "Experiment_names", "ps_16S", "ps_16S_split")

#Remove everything except specified objects
rm(list = setdiff(ls(), keep_objects))

```

#(c)Stress-responsive ASVs
In here we will calculate differentially abundant (DA) ASVs with ANCOMBC as well as their log-fold change respect to control. This information and other metadata will be converted into text files to be used as input for iTOL visualization. 

##Data/Stats
```{r Responsive ASVs; ANCOMBC}

ps_16S_split #original ps object split by Experiment

#Now, run ANCOMBC in each experiment independently using lapply (it will take a while)
ancom_16S <- lapply(ps_16S_split, function(ps){
  set.seed(123)
  ANCOMBC::ancombc(phyloseq = ps, 
                      formula = "plant_pathway_induced", 
                      p_adj_method = "fdr", #same as "BH" Benjamini & Hochberg
                      prv_cut = 0.20, #keep only those ASVs present in at least 20% of the samples
                      group = "plant_pathway_induced", 
                      struc_zero = TRUE, #with structural zero's
                      neg_lb = FALSE, #classifies structural zero's if it's zero and not its asymptotic lower bound 
                      tol = 1e-5,               
                      max_iter = 100, 
                      conserve = TRUE, #conservative variance of the test statistics 
                      alpha = 0.05,
                      verbose=TRUE,
                      global=FALSE)})

#change name to not repeat the code
ancom <- ancom_16S

#get results
head(ancom$cond_phytohormones$feature_table)

#check
lapply(ancom, function(ancom_result){head(ancom_result$res$lfc)})#LFC
lapply(ancom, function(ancom_result){head(ancom_result$res$se)})#standard error
lapply(ancom, function(ancom_result){head(ancom_result$res$W)}) #test statistic LFC/se

#get sample fractions
samp_frac <- lapply(ancom, function(ancom_result){ancom_result$samp_frac})#ancom$samp_frac

# Get log transformed counts and add pseudo-count (1) to feature table to avoid taking the log of 0
log_obs_abn <-lapply(ancom, function(ancom_result){log(ancom_result$feature_table +1)})

# Adjust the log observed abundances with the sample fraction
log_corr_abn <-mapply(function(x,y){
                     t(t(x) - y)},
                     x=log_obs_abn,
                     y=samp_frac)
  
#Results
log_corr_abn #bias-corrected log abundances
lfc <- lapply(ancom, function(ancom_result){ancom_result$res$lfc}) #log-fold change according to control 
se_lfc <- lapply(ancom, function(ancom_result){ancom_result$res$se}) #standard error of log-fold change

#get significant ASVs, keep this to make Venn Diagrams
JA_asvs <- lapply(ancom, function(ancom_result){
           asvs <- ancom_result$res$diff_abn %>%
            dplyr::filter(plant_pathway_inducedJA == TRUE) #26,12,8
           asvs$ASV <- asvs$taxon
           return(asvs)})

SA_asvs <- lapply(ancom, function(ancom_result){
           asvs <- ancom_result$res$diff_abn %>%
            dplyr::filter(plant_pathway_inducedSA == TRUE) #9,4,10
           asvs$ASV <- asvs$taxon
           return(asvs)})

#Now I need to get the unique names 
sig_asvs <- mapply(function(x,y){
  c(x$ASV, y$ASV)},
  x=JA_asvs,
  y=SA_asvs)

#make it into a dataframe format
sig_asvs <- lapply(sig_asvs, function(x){
   unique <- unique(x)
   unique <- as.data.frame(unique)
   colnames(unique) <- "ASV"
   return(unique)})  

#With this I can slice the different data frames to plot them

#1.bias-corrected abundances
log_corr_abn_sig <- mapply(function(x,y){x[y$ASV,]},
  x=log_corr_abn,
  y=sig_asvs) #works because its a matrix 

#2.log-fold change vs controls
lfc_sig <- mapply(function(x,y){x[x$taxon %in% y$ASV,]},
                  x=lfc,
                  y=sig_asvs,
                  SIMPLIFY = FALSE) #to keep it as a list and not a single thing

#3.Differential abundance according to stress 
dif_abun_stress_sig <- mapply(function(x,y){x$res$diff_abn[x$res$diff_abn$taxon %in% y$ASV,]},
                           x=ancom,
                           y=sig_asvs,
                           SIMPLIFY = FALSE) 

#place ASV as rowname to avoid mistakes later
dif_abun_stress_sig <- lapply(dif_abun_stress_sig, function(df){
  rownames(df) <- df$taxon
  df <- df[,3:4]
  return(df)
})

```

```{r Responsive ASVs; Saving results per Experiment}

#---- Phytohormone-treated ----
ancom$cond_phytohormones #ancom results
lfc$cond_phytohormones  #lfc values  
asvs <- sig_asvs$cond_phytohormones$ASV #vector with ASV names 
dif_abun_stress_sig$cond_phytohormones #contains whether TRUE or FALSE for JA and SA 
log_corr_abn_sig$cond_phytohormones #corrected abundances of DA ASVs
#make new ps with only the DA asvs
ps_DA <- phyloseq::prune_taxa(asvs, ps_16S_split$cond_phytohormones) #ps pruned to only those sig in Phyto Experiment.

#make it into a list to handle it easier
phyto_dataDA <- list (ancom_test=ancom$cond_phytohormones, lfc=lfc$cond_phytohormones, sig_asvs=asvs,
                    binary_stress=dif_abun_stress_sig$cond_phytohormones,
                    log_corr_abn=log_corr_abn_sig$cond_phytohormones,
                    ps=ps_DA)

#save it, to not run everything again 
save(phyto_dataDA, file="./R_results/Fig_2_Stress_induced_changes_Bacteria/phyto_dataDA.RData")

#---- Insect-infested ----
ancom$cond_real_insects
lfc$cond_real_insects  
asvs_i <- sig_asvs$cond_real_insects$ASV  
dif_abun_stress_sig$cond_real_insects 
log_corr_abn_sig$cond_real_insects 
#make new ps with only the DA asvs
ps_DA_i <- phyloseq::prune_taxa(asvs_i, ps_16S_split$cond_real_insects) #ps pruned to only those sig in Phyto Experiment.

#make it into a list to handle it easier
insects_dataDA <- list (ancom_test=ancom$cond_real_insects, lfc=lfc$cond_real_insects  , sig_asvs=asvs_i,
                    binary_stress=dif_abun_stress_sig$cond_real_insects,
                    log_corr_abn=log_corr_abn_sig$cond_real_insects,
                    ps=ps_DA_i)

#save it, to not run everything again 
save(insects_dataDA, file="./R_results/Fig_2_Stress_induced_changes_Bacteria/insects_dataDA.RData")


```


##iTOL figure

###Microbial phylogenetic tree
In this case the figure was made with iTOL, and now we are going to make dataframes so they can be imported and visualized in iTOL. 

```{r Responsive ASVs; Sequences for new tree}

#We will need:
ps_16S
phyto_dataDA
insects_dataDA

#check name of ASVs to prune a ps 
p <- rownames(phyto_dataDA$ps@tax_table)
i <- rownames(insects_dataDA$ps@tax_table)

#save ASV vector
da_asvs <- list(phytohormones=p, insects=i)
saveRDS(da_asvs, "./R_results/Fig_2_Stress_induced_changes_Bacteria/da_asvs.rds")

intersect(p,i) #no
asvs <- c(p,i)
asvs <- unique(asvs) #44

#First, let's get the tree by pruning the original ps_16 with DA ASVs
#make ps with DA asvs
ps_DA <- prune_taxa(asvs, ps_16S) #44 taxa, ok!
ps_DA@phy_tree
plot(ps_DA@phy_tree) #this is the default tree

#Now, let's get the sequences to make a new tree as the default one with DADA2 is not great
#get the sequences and save them as FASTA format
ps_DA %>%
refseq() %>%
Biostrings::writeXStringSet("./R_results/Fig_2_Stress_induced_changes_Bacteria/selected_asvs.fna",
             append=FALSE,compress=FALSE,
             compression_level=NA, format="fasta")
```

The annotated tree in iTOL looks great, except the phylogeny seem to be not super good. Thus a recommendation from Chrats Melkonian and Melissa Uribe was to build again the microbial phylogeny. For this, the sequences from the selected_asvs were used to build a new phylogenetic tree using MAFFT, Guidance2 and ModelFinder.  

```{r Responsive ASVs; load new phylogenetic tree}
#Load new phylogenetic tree 
new_tree <- read.tree("./R_results/Fig_2_Stress_induced_changes_Bacteria/16S_DA_tree_IQTREE.tre")
new_tree <- ape::multi2di(new_tree) #root it
plot(new_tree)

#change tree in the ps object
#copy ps 
ps_DA_tree <- ps_DA

#change it for new
ps_DA_tree@phy_tree <- new_tree

#compare it
plot(ps_DA@phy_tree)
plot(ps_DA_tree@phy_tree) #this new tree is rooted 

#save it, this is the one to be used in iTOL
ape::write.tree(new_tree, "./R_results/Fig_2_Stress_induced_changes_Bacteria/16S_DA_tree_IQTREE_rooted.tre")

#call it again
Bac_tree <- ps_DA@phy_tree

#clean
rm(ps_DA_tree)

```

###Full dataframe
Now that the tree is ready, first make a dataframe with all the info needed.
```{r Responsive ASVs; iTOL complete dataframe}

##let's get D1:Experiment
df.p <- as.data.frame(p) %>% 
         mutate(Phytohormones= "phytohormones")%>%
         mutate(Experiment="phytohormones")  
colnames(df.p)[1] <-  "id"
  
df.i <- as.data.frame(i)%>% 
         mutate(Insects= "insects")%>%
         mutate(Experiment="insects") 
colnames(df.i)[1] <-  "id"

#merge
df_experiment <- full_join(df.p, df.i)

#get order
order_asvs <- df_experiment$id

#Now, let's get D2:LFC
#get LFC values from both experiments in a single df 
lfc_phyto <- phyto_dataDA$lfc %>%  mutate(Phytohormones= "phytohormones") 
lfc_phyto <- lfc_phyto[,c(1,3:5)] #remove intercept
colnames(lfc_phyto) <- c("id", "lfc_JA_phyto", "lfc_SA_phyto", "Phytohormones")                              

lfc_insects <- insects_dataDA$lfc %>%  mutate(Insects= "insects")
lfc_insects <- lfc_insects[,c(1,3:5)]
colnames(lfc_insects) <- c("id", "lfc_JA_insects", "lfc_SA_insects", "Insects")  

#let's save also the lfc of only those DA
lfc_phyto_sel <- lfc_phyto %>% filter(id %in% p)
lfc_insect_sel <- lfc_insects %>% filter(id %in% i)

#save LFC
lfc_da_avs <- list(phytohormones=lfc_phyto_sel, insects=lfc_insect_sel)
saveRDS(lfc_da_avs, "./R_results/Fig_2_Stress_induced_changes_Bacteria/lfc_da_avs.rds")

#merge
df_lfc <- full_join(lfc_phyto_sel, lfc_insect_sel)

#Next, let's get D3:Treatment
#phyto
bin_phyto <- phyto_dataDA$binary_stress %>% rownames_to_column(var="id") %>% mutate(Phytohormones="phytohormones")
colnames(bin_phyto) <- c("id","DA_JA_phyto", "DA_SA_phyto", "Phytohormones")
bin_phyto <- bin_phyto %>% mutate(DA_JA_phyto = case_when(DA_JA_phyto == TRUE ~ "YES",
                                DA_JA_phyto == FALSE ~ "NO")) %>% 
                              mutate(DA_SA_phyto = case_when(DA_SA_phyto == TRUE ~ "YES",
                                DA_SA_phyto == FALSE ~ "NO"))
bin_phyto <- bin_phyto %>%
                mutate(DA_phyto = case_when(
                  DA_JA_phyto == "YES" & DA_SA_phyto == "NO" ~ "JA_Only",
                  DA_JA_phyto == "NO" & DA_SA_phyto == "YES" ~ "SA_Only",
                  DA_JA_phyto == "YES" & DA_SA_phyto == "YES" ~ "JA_and_SA")) #works

#insects
bin_insects <- insects_dataDA$binary_stress %>% rownames_to_column(var="id") %>% mutate(Insects="insects")
colnames(bin_insects) <- c("id","DA_JA_insects", "DA_SA_insects", "Insects")
bin_insects <- bin_insects %>% mutate(DA_JA_insects = case_when(DA_JA_insects == TRUE ~ "YES",
                                DA_JA_insects == FALSE ~ "NO")) %>% 
                              mutate(DA_SA_insects = case_when(DA_SA_insects == TRUE ~ "YES",
                                DA_SA_insects == FALSE ~ "NO"))
bin_insects <- bin_insects %>%
                mutate(DA_insects = case_when(
                  DA_JA_insects == "YES" & DA_SA_insects == "NO" ~ "JA_Only",
                  DA_JA_insects == "NO" & DA_SA_insects == "YES" ~ "SA_Only",
                  DA_JA_insects == "YES" & DA_SA_insects == "YES" ~ "JA_and_SA")) #works
  
#merge
binary_stress <- full_join(bin_phyto, bin_insects) #44

#merge to make it complete
df_complete <- full_join(df_lfc, binary_stress)

#add column with Experiment including Phyto, Insects and both
df_complete <- df_complete %>% mutate(Experiment=case_when(
  Phytohormones == "phytohormones" ~ "phytohormones",
  Insects == "insects" ~ "insects"))

#re-order
df_complete <- df_complete %>% relocate(id, Experiment, Phytohormones, Insects)

#Finally, let's get taxonomy
#clean it first
tax_table(ps_DA)[, colnames(tax_table(ps_DA))] <- gsub(tax_table(ps_DA)[, colnames(tax_table(ps_DA))],     pattern = "[a-z]__", replacement = "")
#make it a df 
tax_table <- as.data.frame(ps_DA@tax_table[,c(1:7)]) #select the interesting ones 
tax_table$id <- rownames(tax_table) #add id column for join 

#merge
df_complete2 <- left_join(df_complete, tax_table)

#save df
saveRDS(df_complete2, "./R_results/Fig_2_Stress_induced_changes_Bacteria/df_complete2.rds")

```

###Experiment
Now you can make individual df to be imported into iTOL.
iTOL has a specific format to add data into the tree, it has to be in a .txt file and in a certain way coded. The package itol.toolkit helps to make an annotation file directly in R. Once the annotation file (AF) is ready you can import it into the iTOL annotator to add metadata to your phylogenetic tree.
```{r Responsive ASVs; iTOL Experiment}

#color management
#Set-up colors
color_JA <- "#E69F00"
color_SA <- "#009D64"
color_phyto <- "#336699"
color_insects <- "#996633"

#Built in colors
display.brewer.all() #gives you the palettes from RColorBrewer

#create Annotation files with using the itol.toolkit package

#Experiment: together [not used in final figure]
temp <- df_complete2%>%
        dplyr::select(c(id,Experiment))

#make df for itol
AF1 <- itol.toolkit::create_unit(data = temp, 
                         key = "AF1_Experiment", 
                         type = "DATASET_COLORSTRIP",
                         tree = Bac_tree)

#change colors and parameters
AF1@specific_themes$strip_label$width <- 30 #width of strip
AF1@specific_themes$strip_label$color_branches <- 0 #no color branches
AF1@data$tip <- AF1@data$tip %>%  mutate(`AF1_Experiment$COLOR`=case_when( #change colors 
                                `AF1_Experiment$LABEL`=="phytohormones" ~ color_phyto,
                                `AF1_Experiment$LABEL`=="insects" ~ color_insects))


AF1@common_themes$legend$title <- "Experiment"
AF1@common_themes$legend$horizontal <- 0
AF1@common_themes$legend$labels <- c("Herbivore-induced","Phytohormone-induced")

#save
itol.toolkit::write_unit(AF1, "./R_results/Fig_2_Stress_induced_changes_Bacteria")

```

```{r Responsive ASVs; iTOL Treatment v1}

## Treatment version 1
df_Treatment <- df_complete2%>%
                dplyr::select(id,DA_JA_phyto,DA_JA_insects, DA_SA_phyto, DA_SA_insects)%>%
                relocate(id)
#replace NA's
df_Treatment$DA_JA_phyto <- df_Treatment$DA_JA_phyto %>% replace_na('NO')
df_Treatment$DA_SA_phyto <- df_Treatment$DA_SA_phyto %>% replace_na('NO')
df_Treatment$DA_JA_insects <- df_Treatment$DA_JA_insects %>% replace_na('NO')
df_Treatment$DA_SA_insects <- df_Treatment$DA_SA_insects %>% replace_na('NO')

#replace 0's for -1 and Yes for 1 so iTOL recognizes full and empty symbols
df_Treatment2 <- df_Treatment %>% 
                mutate(DA_JA_phyto=case_when(DA_JA_phyto== "YES" ~ "1", TRUE ~ "-1")) %>% 
                mutate(DA_SA_phyto=case_when(DA_SA_phyto== "YES" ~ "1", TRUE ~ "-1")) %>%
                mutate(DA_JA_insects=case_when(DA_JA_insects== "YES" ~ "1", TRUE ~ "-1")) %>%
                mutate(DA_SA_insects=case_when(DA_SA_insects== "YES" ~ "1", TRUE ~ "-1"))

## By Experiment
#Phyto
Phyto <- df_Treatment2 %>% dplyr::select(id, DA_JA_phyto, DA_SA_phyto)
AF2.1 <- itol.toolkit::create_unit(data = Phyto, 
                         key = "AF2.1_Treatment_Phyto", 
                         type = "DATASET_BINARY",
                         tree = Bac_tree)

AF2.1@field$labels <- c("JA_phyto",  "SA_phyto")
AF2.1@field$colors <- c("#E69F00",  "#009D64")

itol.toolkit::write_unit(AF2.1, "./R_results/Fig_2_Stress_induced_changes_Bacteria")

#Insects
insects <- df_Treatment2 %>% dplyr::select(id, DA_JA_insects, DA_SA_insects)
AF2.2 <- itol.toolkit::create_unit(data = insects, 
                         key = "AF2.2_Treatment_insects", 
                         type = "DATASET_BINARY",
                         tree = Bac_tree)

AF2.2@field$labels <- c("JA_insects",  "SA_insects")
AF2.2@field$colors <- c("#E69F00",  "#009D64")

itol.toolkit::write_unit(AF2.2, "./R_results/Fig_2_Stress_induced_changes_Bacteria")

```

Independently
```{r Responsive ASVs; iTOL Treatment v2}

## Treatment version 2
##JA
JA_phyto <- df_Treatment2 %>% dplyr::select(id, DA_JA_phyto)
AF3.1 <- itol.toolkit::create_unit(data = JA_phyto, 
                         key = "AF3.1_Treatment_JA_Phyto", 
                         type = "DATASET_BINARY",
                         tree = Bac_tree)

AF3.1@field$labels <- "JA_phyto"
AF3.1@field$colors <- "#E69F00"

itol.toolkit::write_unit(AF3.1, "./R_results/Fig_2_Stress_induced_changes_Bacteria")

JA_insects <- df_Treatment2 %>% dplyr::select(id, DA_JA_insects)
AF3.2 <- itol.toolkit::create_unit(data = JA_insects, 
                         key = "AF3.1_Treatment_JA_insects", 
                         type = "DATASET_BINARY",
                         tree = Bac_tree)

AF3.2@field$labels <- "JA_insects"
AF3.2@field$colors <- "#E69F00"

itol.toolkit::write_unit(AF3.2, "./R_results/Fig_2_Stress_induced_changes_Bacteria")

##SA
SA_phyto <- df_Treatment2 %>% dplyr::select(id, DA_SA_phyto)
AF3.3 <- itol.toolkit::create_unit(data = SA_phyto, 
                         key = "AF3.3_Treatment_SA_Phyto", 
                         type = "DATASET_BINARY",
                         tree = Bac_tree)

AF3.3@field$labels <- "SA_phyto"
AF3.3@field$colors <- "#009D64"

itol.toolkit::write_unit(AF3.3, "./R_results/Fig_2_Stress_induced_changes_Bacteria")

SA_insects <- df_Treatment2 %>% dplyr::select(id, DA_SA_insects)
AF3.4 <- itol.toolkit::create_unit(data = SA_insects, 
                         key = "AF3.4_Treatment_SA_insects", 
                         type = "DATASET_BINARY",
                         tree = Bac_tree)

AF3.4@field$labels <- "SA_insects"
AF3.4@field$colors <- "#009D64"

itol.toolkit::write_unit(AF3.4, "./R_results/Fig_2_Stress_induced_changes_Bacteria")

```

### LFC
```{r Responsive ASVs; iTOL LFC}

## LFC dataset
df_LFC <- df_complete2%>%
                dplyr::select(id,lfc_JA_phyto, lfc_JA_insects, lfc_SA_phyto, lfc_SA_insects)%>%
                relocate(id)%>%
                as.tibble()
#let's round up
df_LFC <- df_LFC %>% 
          dplyr::mutate(across(where(is.numeric), round, 2))

#First, lets make a single df 
df_LFC_unique <- full_join(df_Treatment,df_LFC)
#Now let's apply some rules
df_LFC_unique <- df_LFC_unique %>%
                  mutate(lfc_JA_phyto=
                           case_when(DA_JA_phyto == "NO" ~ NA, TRUE ~ lfc_JA_phyto))%>% 
                  mutate(lfc_SA_phyto=
                           case_when(DA_SA_phyto == "NO" ~ NA, TRUE ~ lfc_SA_phyto))%>% 
                  mutate(lfc_JA_insects=
                           case_when(DA_JA_insects == "NO" ~ NA, TRUE ~ lfc_JA_insects))%>% 
                  mutate(lfc_SA_insects=
                           case_when(DA_SA_insects == "NO" ~ NA, TRUE ~ lfc_SA_insects))

#---- Phytohormone-treated ----
Phyto <-  df_LFC_unique %>% dplyr::select(id, lfc_JA_phyto, lfc_SA_phyto)

#get Annotation File for Heatmap
AF4.1 <- itol.toolkit::create_unit(data = Phyto , 
                         key = "AF4.1_LFC_Heatmap_Phyto", 
                         type = "DATASET_HEATMAP",
                         tree = Bac_tree)  #works

AF4.1@data$tip[AF4.1@data$tip == 0] <- NA

#choose colors
min <-"#D6604D"
max <- "#4393C3"

#select colors 
AF4.1@specific_themes$heatmap$color$min <- min
AF4.1@specific_themes$heatmap$color$max <- max
AF4.1@specific_themes$heatmap$color$mid <- "#ebeceb"
AF4.1@specific_themes$heatmap$value$mid <- 0
AF4.1@specific_themes$heatmap$color$nan <- "#FFFFFF"

#change range so the colors mean the same across experiment and treatment
AF4.1@specific_themes$heatmap$value$max <- 2.75
AF4.1@specific_themes$heatmap$value$min <- -2.55
AF4.1@specific_themes$heatmap$value$mid <- 0

AF4.1@specific_themes$heatmap$tree$tree_display <- 0 #no display of tree in heatmap

itol.toolkit::write_unit(AF4.1, "./R_results/Fig_2_Stress_induced_changes_Bacteria") #works

#---- Insect-Infested ----
Insects <-  df_LFC_unique %>% dplyr::select(id, lfc_JA_insects, lfc_SA_insects)

#get Annotation File for Heatmap
AF4.2 <- itol.toolkit::create_unit(data = Insects , 
                         key = "AF4.2_LFC_Heatmap_Insects", 
                         type = "DATASET_HEATMAP",
                         tree = Bac_tree)  #works

AF4.2@data$tip[AF4.2@data$tip == 0] <- NA

#select colors 
AF4.2@specific_themes$heatmap$color$min <- min
AF4.2@specific_themes$heatmap$color$max <- max
AF4.2@specific_themes$heatmap$color$mid <- "#ebeceb"
AF4.2@specific_themes$heatmap$value$mid <- 0
AF4.2@specific_themes$heatmap$color$nan <- "#FFFFFF"

#change range so the colors mean the same across experiment and treatment
AF4.2@specific_themes$heatmap$value$max <- 2.75
AF4.2@specific_themes$heatmap$value$min <- -2.55
AF4.2@specific_themes$heatmap$value$mid <- 0

AF4.2@specific_themes$heatmap$tree$tree_display <- 0 #no display of tree in heatmap

itol.toolkit::write_unit(AF4.2, "./R_results/Fig_2_Stress_induced_changes_Bacteria") #works

```

###Taxonomy
```{r Responsive ASVs; iTOL LFC}

#first, let's change it for Family labels [not used in final Figure]
#Family
df_labels_F <- df_complete2 %>%
               dplyr::select(id, Family)

AF5 <- create_unit(data = df_labels_F,
                       key = "AF5_label_Family",
                       type = "LABELS",
                       tree = Bac_tree)

itol.toolkit::write_unit(AF5, "./R_results/Fig_2_Stress_induced_changes_Bacteria")


#Genus
df_labels_G <- df_complete2 %>%
               dplyr::select(id, Genus)

#replace Genus names with names of ASV and add number to Genus so its clear that it's different ASVs
df_labels_G <- df_labels_G %>% mutate(number=str_split_i(df_labels_G$id, "_", 2))%>% #get number of ASV
                                mutate(Genus=paste(.$Genus, .$number, sep= "-"))%>% #make new with number and genus
                                mutate(Genus=gsub("NA", "fASV", .$Genus))%>%
                                dplyr::select(-number) #ok
                                      
AF6 <- create_unit(data = df_labels_G,
                       key = "AF6_label_Genus",
                       type = "LABELS",
                       tree = Bac_tree)

itol.toolkit::write_unit(AF6, "./R_results/Fig_2_Stress_induced_changes_Bacteria")


#Make a range df for Phylum now
df_Phylum <- df_complete2 %>%
               dplyr::select(id, Phylum, Class)

#make special category for Proteobacteria to include Class
df_Phylum <- df_Phylum %>% mutate(Phylum=case_when(
                          Class==" Alphaproteobacteria" ~ " Alphaproteobacteria",
                          Class == " Gammaproteobacteria" ~ " Gammaproteobacteria",
                          TRUE ~ Phylum))

#make a vector with colors 
Phylum <- unique(df_Phylum$Phylum) #11
Phylum <- sort(Phylum)
color_phylum <- c("#FDB462","#8DD3C7","#FFFFB3", "#FB8072","#80B1D3","#8DD3D2",
                  "#FCCDE5", "#BC80BD","#BEBADA", "#B3DE69", "#CCEBC5")

#make it a df
df_phyla_col <- data.frame(Phylum, color_phylum)

#merge with df_Phylum
df_Phylum <- df_Phylum%>%
             dplyr::select(-Class) %>% 
             dplyr::inner_join(df_phyla_col)

AF7 <- create_unit(data = df_Phylum, 
                      key = "AF7_Phylum_colors", 
                      type = "TREE_COLORS", 
                      subtype = "range",
                      tree = Bac_tree)

itol.toolkit::write_unit(AF7, "./R_results/Fig_2_Stress_induced_changes_Bacteria")

```

Now, import the tree and the df's into iTOL to visualize it.

#(d)Venn-Diagrams

Proportional Venn Diagrams were made with the online tool from the nVenn package (https://degradome.uniovi.es/cgi-bin/nVenn/nvenn.cgi) for this you just need to copy-paste the list of ASVs responsive to JA or SA in each experiment and download the .svg figure.

##Data
```{r venn diagrams}

#----Phytohormone-treated ----
JA_asvs$cond_phytohormones$ASV #Group 1
SA_asvs$cond_phytohormones$ASV #Group 2

#---- Herbivore-infested ----
JA_asvs$cond_real_insects$ASV #Group 3
SA_asvs$cond_real_insects$ASV #Group 4

```






