---
title: "Untitled"
author: "Marcela Aragon"
date: "2025-02-03"
output: html_document
---

# General settings

## Loading packages

```{r loading packages, echo=TRUE}

#Set working directory to project directory 
setwd("./")
getwd() #ok

#data
library(knitr)
library(dplyr)
library(tibble)

#plotting
library(ggplot2)
library(RColorBrewer)
library(grDevices)
library(rempsyc) #tables

#stats
library(DHARMa)
library(lmtest)
library(fitdistrplus)
library(AICcmodavg)
library(emmeans)
library(multcompView)
library(multcomp)
library(postHoc)
library(car)
library(lsr)

```

## Global settings for plots 

```{r settings for plots}

# Setting up white background for plots
theme_set(theme_bw())

# Set colors for soil_inocula
color_pathway_induced <- c("#989898", "#E69F00", "#009D64") #control, JA and SA
color_experiment <- c("#336699", "#996633") #phytohormone-treated and herbivore-infested

# Set default axis looks
axis_looks <- theme(axis.text.x = element_text(colour = "black", size = 9,
                                               face = "bold", angle=0, hjust=0.5))+ #text in X
  theme(axis.text.y = element_text(colour = "black", size = 9, face = "bold"))+ #text in y
  theme(axis.title=element_text(size=11, face = "bold"))+ #all texts
  theme(axis.title.y = element_text(margin = ggplot2::margin(t = 0, r = 10, b = 0, l = 0)))+
  theme(panel.grid.major = element_blank(), #no grid
        panel.grid.minor = element_blank()) #no grid
  #theme(legend.position="none")+ #no legend

#Get names for boxes in facet wrap
Experiment_names <- c(
                    `simulated` = "Phytohormone-treated",
                    `real_insects` = "Herbivore-infested")

```



# (a)Shoot biomass in conditioning phase 

## Data
Load and fix biomass data for conditioning phase
```{r shoot_biomass conditioning data}

#load data
biomass <- readRDS("./Data/plant_data/biomass.rds")

#quick summary
biomass %>% 
group_by(Phase, Experiment, plant_pathway_induced) %>%
  summarise(
    count = n(),
    mean = mean(shoot_biomass_mg, na.rm = TRUE),
    sd = sd(shoot_biomass_mg, na.rm = TRUE))


#subsetting df to only conditioning
conditioning <- biomass %>% 
                filter(Phase == "conditioning")


#change levels of Experiment; simulated=phytohormone-treated, real_insects=herbivore-infested
levels(conditioning$Experiment)
conditioning <- conditioning %>%
                mutate(Experiment = factor(Experiment, levels = c("simulated", "real_insects" )))

```

## Plot
Plot it
```{r shoot_biomass conditioning plot}

# make a boxplot
p1 <- ggplot(data = conditioning,
            mapping = aes(x = plant_pathway_induced, y = shoot_biomass_mg), colour= plant_pathway_induced)+
  geom_boxplot(mapping = aes(fill= plant_pathway_induced), position = position_dodge(width = 0.8), lwd=0.8)+
  geom_jitter(data = conditioning,
              mapping = (aes(plant_pathway_induced, shoot_biomass_mg, fill = plant_pathway_induced)),
              shape = 21, size = 1.8, color = "black",
              position = position_jitterdodge(jitter.height = .1, jitter.width = .2))+
  scale_fill_manual(values = color_pathway_induced)

#facet_wrap it by experiment and add info
p2 <- p1 + facet_wrap(~Experiment, labeller = as_labeller(Experiment_names)) +
              labs(y="Shoot biomass (mg)", x=" ") +
              ylim(400,2500)+
              axis_looks +
              theme(axis.text.x = element_blank(),
                    axis.ticks.x = element_blank())+
              theme(strip.text.x = element_text(size = 11, face = "bold"), #text in boxes
                    strip.background=element_rect(colour="black"))+ #outline of boxes
              theme(panel.spacing = unit(0.2, "lines"))+ #space in between facets
              theme(legend.position="none") #no legend, to save it better
#check plot
p2


```

## Stats
Run stats split by experiment, first phytohormone-treated
```{r shoot_biomass conditioning stats, phytohormone-treated}

#subset to phyto
phyto <- subset(conditioning, Experiment == "simulated") %>% droplevels()

#quick summary
phyto %>% group_by(Experiment, plant_pathway_induced) %>% summarise(n=n()) #balanced

#run glm with normal distribution
aov_simulated <- glm(shoot_biomass_mg  ~ plant_pathway_induced,
                     family=gaussian,
                     data=phyto)
#get summary
lm.anova <- car::Anova(aov_simulated, type="III", test.statistic="LR")

#check model assumptions
simulateResiduals(aov_simulated, plot=T) #ok

#post-hoc test
#pairiwse differences
emmeans(aov_simulated, pairwise~plant_pathway_induced) 

```


```{r shoot_biomass conditioning stats, herbivore-infested}

#subset to phyto
herb <- subset(conditioning, Experiment == "real_insects") %>% droplevels()

#quick summary
herb %>% group_by(Experiment, plant_pathway_induced) %>% summarise(n=n()) #balanced

#run glm with normal distribution
aov_insects <- glm(shoot_biomass_mg  ~ plant_pathway_induced,
                     family=gaussian,
                     data=herb)
#get summary
lm.anova.i <- car::Anova(aov_insects, type="III", test.statistic="LR")

#check model assumptions
simulateResiduals(aov_insects, plot=T) #ok

#post-hoc test
#pairiwse differences
emmeans(aov_insects, pairwise~plant_pathway_induced) 

```

# (b)PLS-DA endogenous phytohormones

Stands for Partial Least Squares Discriminant Analysis; a type of multivariate dimensionality reduction analysis. It can be thought as a "supervised" version of a PCA. The general approach is to analyze all phytohormones in a single analysis to check for patterns.  

https://omicsforum.ca/t/what-is-the-meaning-of-components-in-pls-da/732/3

## Data
```{r PLS-DA phytohormones conditioning data}

#load data
phytohormones <- readRDS("./Data/plant_data/phytohormones.rds")

#change levels of Experiment
phyto <- phytohormones %>%
                mutate(Experiment = factor(Experiment,
                                           levels = c("simulated", "real_insects" )))%>%
                mutate(plant_pathway_induced = factor(plant_pathway_induced,
                                                      levels=c("Control", "JA", "SA")))

#subsetting df to only conditioning
conditioning <- phytohormones %>% 
                      filter(Phase == "conditioning")

#make long format for plotting
long <- phyto %>% 
              tidyr::pivot_longer(cols = c(11:18), 
              names_to = "phytohormone",
              values_to = "concentration",
              values_drop_na = TRUE)

#making new phytohormone column as factor
long$phytohormone <- as.factor(phyto_long$phytohormone)

# Reordering downstream the JA pathway
ordered_hormones <- c("ABA", "SA", "cis_OPDA", "JA", "OH_JA", "sum_JA_Ile", "OH_JA_Ile", "COOH_JA_Ile")

long$phytohormone <-factor(long$phytohormone, 
                             levels = ordered_hormones)

#subsetting df to only conditioning
conditioning_long <- long %>% 
                filter(Phase == "conditioning")

#Now I have 3 df, phyto, long and conditioning 

```

Now let's calculate the PLS-DA
```{r PLS-DA phytohormones conditioning data2}
#Some of the following code was written by Els van de Zande, thank you Els!

#load libraries
library(ropls) #PLS-DA
library(mixOmics) #PLS-DA

#get matrix
matrixdata<-conditioning[,11:18]# keep only values
matrixdata <- as.matrix(matrixdata[,])# make it numeric
metadata<-conditioning[,c("Experiment", "plant_pathway_induced")]
metadata <- metadata %>% mutate(joined=paste(Experiment, plant_pathway_induced, sep="_"))
metadata <- factor(metadata$joined)

#PLS-DA 
plsda <-opls(matrixdata, metadata, scaleC = "pareto", permI = 100)
#pareto removes the relative importance of large values
#but keeps the data structure intact.

#get summary
plot(plsda, typeVc = "x-score", parCexN = 1.5)
vip_scores <- getVipVn(plsda)

#let's remove those with VIP < 0.75
low_vip <- which(vip_scores < 0.75)
filtered_matrixdata <- matrixdata[,-low_vip]

#re-run PLSDA, good one
plsda_filt <-opls(filtered_matrixdata, metadata, scaleC = "pareto", permI = 100)
plot(plsda_filt)

#get scores
scores<-plsda_filt@scoreMN
scores_df<-as.data.frame(scores) #make new df 

scores_df$p1 <- as.numeric(scores_df$p1)
scores_df$p2 <- as.numeric(scores_df$p2)

metadata<-conditioning[,c("Experiment", "plant_pathway_induced")]

scores_df2 <- scores_df %>% #add info
              mutate(Experiment = as.numeric(metadata$Experiment))%>%
              mutate(plant_pathway_induced = as.numeric(metadata$plant_pathway_induced))
```

## Plot
Now make the bi-plot with scores and loadings
```{r PLS-DA phytohormones conditioning plot}

#Plot the scores using ggplot2
scoreplot<-ggplot(scores_df2, aes(x = p1, y = p2, fill = factor(plant_pathway_induced),
                                  shape = factor(Experiment))) +
  geom_point(aes(colour=factor(plant_pathway_induced), fill=factor(plant_pathway_induced)),
              size = 4, alpha=0.8)+
  labs(x = "t1(69%)", y = "t2(17%)",
       color="Plant pathway induced", shape="Experiment", fill=element_blank()) +
  scale_fill_manual(values =color_pathway_induced) +  #Customize color scheme for Treatment
  scale_color_manual(values =color_pathway_induced)
  scale_shape_manual(values = c(16,17)) +  #Customize shape for Herbivore
  axis_looks

#now add loadings
loadings<-plsda_filt@loadingMN
loadings_df<-as.data.frame(loadings)

loadings_df$p1 <- as.numeric(loadings_df$p1)
loadings_df$p2 <- as.numeric(loadings_df$p2)

#makes plot with arrows marking direction
hormone_names <- c("SA","JA","ABA","cis-OPDA","OH-JA")

Biplot <-scoreplot+
         geom_hline(yintercept=0, linetype='dotted', col = 'gray20',linewidth = 0.2 )+
         geom_vline(xintercept=0, linetype='dotted', col = 'gray20', linewidth = 0.2)+
         geom_segment(data = loadings_df, aes(x = 0, y = 0, xend = (p1*50), yend = (p2*50)),
                      arrow = arrow(length = unit(0.5, "picas"), type="closed"),
                      colour = "black", linewidth = 0.5, inherit.aes = FALSE)+
         geom_text(aes(x = (p1*50), y = (p2*50), label = hormone_names,
                angle = 0.45, hjust = -.1, vjust = .2),
                 data = loadings_df,
                 colour = "black", size=4, inherit.aes = FALSE)+
        axis_looks

Biplot

```

## Stats
Analyze the data with PERMANOVA. First, data is auto-scaled to calculate a distance matrix from which a PERMANOVA is run to see whether there is an effect of plant pathway induced, the type of induction and the interaction of these two factors
```{r PLS-DA phytohormones conditioning stats}

library(vegan)

#PERMANOVA
UVmatrixdata<- santaR:::scaling_UV(filtered_matrixdata) #for auto-scaling using st. deviation
DistanceMatrix<-dist(UVmatrixdata, method="euclidean")

#make it a df again
MetaData <-as.data.frame(metadata) 
MetaData[,c(1:2)]<- lapply(MetaData[,c(1:2)], as.factor)

#run PERMANOVA
permanova <- adonis2(DistanceMatrix~Experiment*plant_pathway_induced,
                      data=MetaData, permutations = 999)

#check results
permanova

```

# (c)GLS content in conditioning phase 
##Data
```{r GLS conditioning data}

#load data
gluco <- readRDS("./Data/plant_data/glucosinolates.rds")

#change levels of Experiment; simulated=phytohormone-treated and real_insects=herbivore-infested.
levels(gluco$Experiment)
gluco <- gluco %>%
                mutate(Experiment = factor(Experiment,
                                           levels = c("simulated", "real_insects" )))%>%
                mutate(plant_pathway_induced = factor(plant_pathway_induced,
                                                      levels=c("Control", "JA", "SA")))
#remove sample
gluco <-gluco %>%
        subset(ID != "159")%>% #dry sample, no info
        droplevels()

#### classify by Indolic or Aliphatic ####
indolic <- c("Glu_4OHI3M", "Glu_I3M", "Glu_1MOI3M", "Glu_4MOI3M") 
aliphatic <- c("Glu_Allyl", "Glu_3MSOP","Glu_4MSOB")
#make a small df
values <- c(indolic, aliphatic)
#Create a vector of names, repeating the vector names accordingly
names <- c(rep("indolic", length(indolic)), rep("aliphatic", length(aliphatic)))
# Create the data frame
key_glucosinolates <- data.frame(type = names, glucosinolate = values)
#clean
rm(values,names)
#### ####

#order it
gluco <- gluco %>%
          relocate(Glu_Allyl,Glu_3MSOP,Glu_4MSOB, .after=weight_mg) %>% 
          relocate(Glu_4OHI3M,Glu_I3M,Glu_1MOI3M, Glu_4MOI3M, .after=Glu_4MSOB)%>% 
          mutate(Glu_alyphatic=rowSums(.[11:13]))%>%
          mutate(Glu_indolic=rowSums(.[14:17]))


#make long format for plotting
long <- gluco %>% 
              tidyr::pivot_longer(cols = c(11:20), 
              names_to = "glucosinolate",
              values_to = "concentration",
              values_drop_na = TRUE)

#making new glucosinolate column as factor
long$glucosinolate <- as.factor(long$glucosinolate)

#subsetting df to only conditioning
conditioning_long <- long %>% 
                filter(Phase == "conditioning")%>%
                filter(glucosinolate != "Glu_4OHI3M") #mostly 0's, remove from analysis 

conditioning_gluco <- gluco %>%
                      dplyr::filter(Phase == "conditioning")%>%
                      dplyr::select(-ends_with('4OHI3M'))#mostly 0's, remove from analysis

#first filter by aliphatic and indolic glucosinolates
long_cond <- conditioning_long %>% filter(
  glucosinolate %in% c("Glu_alyphatic","Glu_indolic")&
    Phase == "conditioning")

#Now make a summary df
df_GLS_p <- long_cond %>% 
            group_by(Experiment, plant_pathway_induced, glucosinolate) %>%
  summarise(
    n = n(),
    mean = mean(concentration, na.rm = TRUE),
    sd = sd(concentration, na.rm = TRUE)) %>%
    mutate(se = sd/sqrt(n))

#Now I have 3 df, gluco, conditioning_long and conditioning_gluco 

```

##Plot
```{r GLS conditioning plot}

#Finally, plot it
p3 <- ggplot(data = df_GLS_p,
       mapping = aes(x = glucosinolate, y = mean), colour= plant_pathway_induced)+
  geom_bar(aes(fill = plant_pathway_induced), stat = "identity",
           position = position_dodge(0.8), width = 0.7, color = "black")+
  geom_errorbar(aes(ymin = mean-se, ymax = mean+se, group = plant_pathway_induced),
                width = 0.2, position = position_dodge(0.8))+
  geom_jitter(data = long_cond,
              mapping = (aes(glucosinolate, concentration,
                             fill = plant_pathway_induced)),
              shape = 21, size = 1.5, color = "black",
              position = position_jitterdodge(jitter.height = .1, jitter.width = .2))+
  scale_fill_manual(values = color_pathway_induced)+
  axis_looks+
  facet_wrap(~Experiment)

#check it
p3

```

##Stats
Let's analyze it independently by experiment and GLS
```{r GLS conditioning stats; Phythormone-treated}

#Phytohormone-treated
## Aliphatic 
phyto.a <- subset(long_cond, Experiment == "simulated" & glucosinolate == "Glu_alyphatic")

phyto.a %>% group_by(Experiment, plant_pathway_induced) %>% summarise(n=n()) #balanced N=4

#run glm
lm <- glm(log(concentration) ~ plant_pathway_induced,
          family=gaussian,
          data=phyto.a)

#get summary 
lm_table <- car::Anova(lm, type="II", test.statistic="LR") #no effect of stress treatment

#check model assumptions
simulateResiduals(lm, plot=T) #ok

##Indole
phyto.i <- subset(long_cond, Experiment == "simulated" & glucosinolate == "Glu_indolic")

phyto.i %>% group_by(Experiment, plant_pathway_induced) %>% summarise(n=n()) #balanced N=4

#run glm
lm <- glm(log(concentration) ~ plant_pathway_induced,
          family=gaussian,
          data=phyto.i)

#get summary 
lm_table <- car::Anova(lm, type="II", test.statistic="LR") #sig effect of stress treatment

#check model assumptions
simulateResiduals(lm, plot=T) #ok

#post-hoc
#emmeans
emmeans <- emmeans(lm, pairwise~plant_pathway_induced)
cld(emmeans, Letters="letters", adjust="tukey")

```

```{r GLS conditioning stats; Herbivore-infested}

#Herbivore-infested
##Aliphatic 
insects.a <- subset(long_cond, Experiment == "real_insects" & glucosinolate == "Glu_alyphatic")
insects.a %>% group_by(plant_pathway_induced) %>% summarise(n=n()) 

#run glm
lm <- glm(log(concentration) ~ plant_pathway_induced,
          family=gaussian,
         data=insects.a)

#summary
lm_table <- car::Anova(lm, type="III", test.statistic="LR")

#check model assumptions
simulateResiduals(lm, plot=T)

#post-hoc
#emmeans
emmeans <- emmeans(lm, pairwise~plant_pathway_induced)
cld(emmeans, Letters="letters", adjust="tukey")

##Indole
insects.i <- subset(long_cond, Experiment == "real_insects" & glucosinolate == "Glu_indolic")
insects.i %>% group_by(plant_pathway_induced) %>% summarise(n=n()) 

#run glm
lm <- glm(log(concentration) ~ plant_pathway_induced,
          family=gaussian,
         data=insects.i)

#summary
lm_table <- car::Anova(lm, type="III", test.statistic="LR")

#check model assumptions
simulateResiduals(lm, plot=T)

#post-hoc
#emmeans
emmeans <- emmeans(lm, pairwise~plant_pathway_induced)
cld(emmeans, Letters="letters", adjust="tukey")

```
