---
title: "Figure 4: Plant responses in feedback phase"
author: "Marcela Aragon"
date: "2025-02-03"
output: html_document
---

# General settings

## Loading packages

```{r loading packages}

#Set working directory to project directory 
setwd("./")
getwd() #ok

#data
library(knitr)
library(dplyr)
library(tibble)
library(forcats)

#plotting
library(ggplot2)
library(RColorBrewer)
library(grDevices)
library(rempsyc) #tables

#stats
library(DHARMa)
library(lmtest)
library(fitdistrplus)
library(AICcmodavg)
library(emmeans)
library(multcompView)
library(multcomp)
library(postHoc)
library(car)
library(lsr)
library(ropls) #PLS-DA
library(mixOmics) #PLS-DA
library(vegan) #PERMANOVA

```

## Global settings for plots 

```{r settings for plots}

# Setting up white background for plots
theme_set(theme_bw())

# Set colors for soil_inocula
color_pathway_induced <- c("#989898", "#E69F00", "#009D64") #control, JA and SA
color_experiment <- c("#336699", "#996633") #phytohormone-treated and herbivore-infested

# Set default axis looks
axis_looks <- theme(axis.text.x = element_text(colour = "black", size = 9,
                                               face = "bold", angle=0, hjust=0.5))+ #text in X
  theme(axis.text.y = element_text(colour = "black", size = 9, face = "bold"))+ #text in y
  theme(axis.title=element_text(size=11, face = "bold"))+ #all texts
  theme(axis.title.y = element_text(margin = ggplot2::margin(t = 0, r = 10, b = 0, l = 0)))+
  theme(panel.grid.major = element_blank(), #no grid
        panel.grid.minor = element_blank()) #no grid
  #theme(legend.position="none")+ #no legend

#Get names for boxes in facet wrap
Experiment_names <- c(
                    `simulated` = "Phytohormone-treated",
                    `real_insects` = "Herbivore-infested")

```

# (a)Shoot biomass in feedback phase 

## Data
Load and fix biomass data for feedback phase

```{r shoot_biomass feedback; data}

#load data
biomass <- readRDS("./Data/plant_data/biomass.rds")

#subsetting df to only feedback
feedback <- biomass %>% 
                filter(Phase == "feedback")


#change levels of Experiment & Herbivory
levels(feedback$Experiment)
levels(feedback$herbivory)
levels(feedback$soil_inocula)

feedback <- feedback %>%
                mutate(Experiment = factor(Experiment, levels = c("simulated", "real_insects" )))%>%
                mutate(herbivory = factor(herbivory, levels = c("Without", "Chewer", "Aphid")))

feedback <- feedback %>%
                mutate(herbivory = factor(herbivory, labels = c("uninfested", "Mamestra", "Myzus")))%>%
                filter(!is.na(shoot_biomass_mg))%>% #remove NAs
                droplevels()

#finally, delete not informative columns
feedback <- feedback %>% dplyr::select(-Phase, -plant_pathway_induced, -Stress, -Leaf_area)

```

##Plot

```{r shoot_biomass feedback; plot}

# make a boxplot for phytohormone-treated experiment
biomass_p <- ggplot(data = subset(feedback, Experiment == "simulated"),
       mapping = aes(x = soil_inocula, y = shoot_biomass_mg), colour= soil_inocula)+
  geom_boxplot(mapping = aes(fill= soil_inocula), position = position_dodge(width = 0.8), lwd=0.8)+
  geom_jitter(data = subset(feedback, Experiment == "simulated"),
              mapping = (aes(soil_inocula, shoot_biomass_mg, fill = soil_inocula)),
              shape = 21, size = 1.8, alpha=0.9, color = "black",
              position = position_jitterdodge(jitter.height = .1, jitter.width = .2))+
  scale_fill_manual(values = color_pathway_induced)+
  axis_looks+
  ylim(100,2000)+
  theme(legend.position="none")+
  facet_wrap(~herbivory)+
  theme(panel.spacing = unit(0, "lines"))

#make a boxpot for insect-infested experiment
biomass_i <- ggplot(data = subset(feedback, Experiment == "real_insects"),
       mapping = aes(x = soil_inocula, y = shoot_biomass_mg), colour= soil_inocula)+
  geom_boxplot(mapping = aes(fill= soil_inocula), position = position_dodge(width = 0.8), lwd=0.8)+
  geom_jitter(data = subset(feedback, Experiment == "real_insects"),
              mapping = (aes(soil_inocula, shoot_biomass_mg, fill = soil_inocula)),
              shape = 21, size = 1.8, alpha=0.9,color = "black",
              position = position_jitterdodge(jitter.height = .1, jitter.width = .2))+
  scale_fill_manual(values = color_pathway_induced)+
  axis_looks+
  ylim(100,2000)+
  theme(legend.position="none")+
  facet_wrap(~herbivory)+
  theme(panel.spacing = unit(0, "lines"))


# Combine the two plots
library(patchwork)
combined_plot <- biomass_p + biomass_i

#check
combined_plot

```


##Stats
```{r shoot_biomass feedback; stats}

#---- Phytohormone-treated ----
phyto <- subset(feedback, Experiment == "simulated")

#quick summary
phyto %>% group_by(Experiment, soil_inocula, herbivory) %>% summarise(n=n()) #not balanced

#run glm with interaction
lm <- glm(shoot_biomass_mg ~ soil_inocula*herbivory,
          family=gaussian,
          data=phyto)

#get summary 
lm.anova <- car::Anova(lm, type="III", test.statistic="LR") #no effect of either

#check assumptions
simulateResiduals(lm, plot=T) #ok

#---- Insect-infested ----
insects <- subset(feedback, Experiment == "real_insects")
insects %>% group_by(Experiment, soil_inocula, herbivory) %>% summarise(n=n()) #balanced

#run glm with interaction
lm.i <- glm(shoot_biomass_mg ~ soil_inocula*herbivory,
            family=gaussian,
            data=insects)

#get summary
lm.i.anova <- car::Anova(lm.i, type="II", test.statistic="LR") #marginal but not sig effect of herbivory

#check assumptions
simulateResiduals(lm.i, plot=T) #ok

```

#(b)PLS-DA endogenous phytohormones
##Data
```{r PLS-DA phytohormones feedback; data}

#load data
phytohormones <- readRDS("./Data/plant_data/phytohormones.rds")

#change levels of Experiment
phyto <- phytohormones %>%
                mutate(Experiment = factor(Experiment,
                                           levels = c("simulated", "real_insects" )))%>%
                mutate(plant_pathway_induced = factor(plant_pathway_induced,
                                                      levels=c("Control", "JA", "SA")))

#make long format for plotting
long <- phyto %>% 
              tidyr::pivot_longer(cols = c(11:18), 
              names_to = "phytohormone",
              values_to = "concentration",
              values_drop_na = TRUE)

#making new phytohormone column as factor
long$phytohormone <- as.factor(phyto_long$phytohormone)

# Reordering downstream the JA pathway
ordered_hormones <- c("ABA", "SA", "cis_OPDA", "JA", "OH_JA", "sum_JA_Ile", "OH_JA_Ile", "COOH_JA_Ile")
long$phytohormone <-factor(long$phytohormone, 
                             levels = ordered_hormones)

#subsetting df to only feedback
feedback <- phytohormones %>% 
                      filter(Phase == "feedback")

#removing samples with NA values 
feedback <- feedback %>% 
            subset(ID != 140 & ID!= 146 & ID!= 159 & ID!= 163 & ID!= 208 & ID!= 215
                       & ID!= 220 & ID!= 222 & ID!= 224)%>% #removing those  with NA values
                droplevels()

#subset long df to only feedback
feedback_long <- long %>% 
                filter(Phase == "feedback")%>%
                subset(ID != 140 & ID!= 146 & ID!= 159 & ID!= 163 & ID!= 208 & ID!= 215
                       & ID!= 220 & ID!= 222 & ID!= 224)%>%
                droplevels()

#change names of soil_inocula (soil treatment)
levels(feedback$soil_inocula)
feedback <- feedback  %>%
            mutate(soil_inocula=factor(soil_inocula, labels=c("control_inocula", 
                                                              "JA_inocula",
                                                              "SA_inocula")))
levels(feedback_long$soil_inocula)
feedback_long <- feedback_long  %>%
                  mutate(soil_inocula=factor(soil_inocula, labels=c("control_inocula", 
                                                              "JA_inocula",
                                                              "SA_inocula")))
#change names of herbivory
levels(feedback$herbivory)
feedback <- feedback  %>%
            mutate(herbivory=factor(herbivory, levels=c("without", "caterpillar", "aphid")))

feedback <- feedback  %>%
            mutate(herbivory=factor(herbivory,
                                    labels=c("uninfested", "Mamestra", "Myzus")))

levels(feedback_long$herbivory)
feedback_long <- feedback_long  %>%
                  mutate(herbivory=factor(herbivory, levels=c("without", "caterpillar", "aphid")))

feedback_long <- feedback_long  %>%
                mutate(herbivory=factor(herbivory,
                                    labels=c("uninfested", "Mamestra", "Myzus")))

#I have feedback and feedback_long df's now

```

Now let's calculate the PLS-DA
```{r PLS-DA phytohormones feedback; data PLSDA}

#Some of the following code was written by Els van de Zande, thank you Els!

#get matrix
matrixdata<-feedback[,11:18]# keep only values
matrixdata <- as.matrix(matrixdata[,])# make it numeric
metadata <-feedback[,c("Experiment", "soil_inocula", "herbivory")]
#metadata <- metadata %>% mutate(joined=paste(Experiment, herbivory, sep="_"))
metadata <- factor(metadata$herbivory) #using only herbivory as explanatory variable

#PLS-DA 
plsda <-opls(matrixdata, metadata, scaleC = "pareto", permI = 100) #herbivory alone is a much better predictor 
#pareto removes the relative importance of large values
#but keeps the data structure intact.

#get summary
plot(plsda, typeVc = "x-score", parCexN = 1.5)
vip_scores <- getVipVn(plsda)

#let's remove those with VIP < 0.75
low_vip <- which(vip_scores < 0.75)
filtered_matrixdata <- matrixdata[,-low_vip]

#re-run PLSDA
plsda_filt <-opls(filtered_matrixdata, metadata, scaleC = "pareto", permI = 100)
plot(plsda_filt)

#get scores
scores<-plsda_filt@scoreMN
scores_df<-as.data.frame(scores) #make new df 

scores_df$p1 <- as.numeric(scores_df$p1)
scores_df$p2 <- as.numeric(scores_df$p2)

metadata<-feedback[,c("Experiment", "soil_inocula", "herbivory")]

scores_df2 <- scores_df %>% #add info
              mutate(Experiment = as.factor(metadata$Experiment))%>%
              mutate(soil_inocula = as.factor(metadata$soil_inocula))%>%
              mutate(herbivory=as.factor(metadata$herbivory))
```

##Plot
Now make the bi-plot with scores and loadings
```{r PLS-DA phytohormones feedback; plot}

str(scores_df2) #ok
ggpubr::show_point_shapes() #check

#plot
scoreplot<-ggplot(scores_df2, aes(x = p1, y = p2,
                                  fill=herbivory,#inside
                                  shape=soil_inocula, #shape
                                  color=Experiment))+ #outline
  geom_point(size = 4, stroke=0.5, alpha=0.8,
             aes(color=Experiment, fill=herbivory))+
  scale_shape_manual(values=c(21,22,24))+
  scale_color_manual(values = c("black", "white"))+ #outline
  scale_fill_manual(values=color_pathway_induced)+ #inside
  labs(x = "LV1 (60%)", y = "LV2 (21%)",
  color="Experiment", shape="herbivory", fill="Soil inocula")+
  axis_looks

#now add loadings
loadings<-plsda_filt@loadingMN
loadings_df<-as.data.frame(loadings)

loadings_df$p1 <- as.numeric(loadings_df$p1)
loadings_df$p2 <- as.numeric(loadings_df$p2)

#makes plot with arrows marking direction
hormone_names <- c("SA","JA","cis-OPDA","OH-JA")

#make biplot
Biplot <-scoreplot+
         geom_hline(yintercept=0, linetype='dotted', col = 'gray20',linewidth = 0.2 )+
         geom_vline(xintercept=0, linetype='dotted', col = 'gray20', linewidth = 0.2)+
         geom_segment(data = loadings_df, aes(x = 0, y = 0, xend = (p1*70), yend = (p2*70)),
                      arrow = arrow(length = unit(0.5, "picas"), type="closed"),
                      colour = "black", linewidth = 0.5, inherit.aes = FALSE)+
         geom_text(aes(x = (p1*70), y = (p2*70), label = hormone_names,
                angle = 0.45, hjust = -.1, vjust = .2),
                 data = loadings_df,
                 colour = "black", size=4, inherit.aes = FALSE)+
        axis_looks

#check
Biplot #legend is wrong, plot is ok

```

##Stats
Now, let's analyze the data with a PERMANOVA. First, data is auto-scaled to calculate a distance matrix from which a PERMANOVA is run to see whether there is an effect of plant pathway induced, the type of induction and the interaction of these two factors. 

```{r PLS-DA phytohormones conditioning stats}

#PERMANOVA
UVmatrixdata<- santaR:::scaling_UV(filtered_matrixdata) #for auto-scaling using st. deviation
DistanceMatrix<-dist(UVmatrixdata, method="euclidean")

#make it a df again
MetaData <-as.data.frame(metadata) 
MetaData[,c(1:3)]<- lapply(MetaData[,c(1:3)], as.factor)

#run PERMANOVA
permanova <- adonis2(DistanceMatrix~Experiment+soil_inocula+herbivory,
                      data=MetaData, permutations = 999) 

#removed interactions as it makes it more difficult to interpret biologically, and R2 values from interactions are very small.

```

#(c)Caterpillar weight
##Data
```{r}

#load data
caterpillars <- readRDS("./Data/plant_data/caterpillars.rds")

#check levels
levels(caterpillars$Experiment) #ok
levels(caterpillars$soil_inocula) #ok

#make it long
cat_long <- caterpillars %>% 
               tidyr::pivot_longer(cols = c(7:11), 
               names_to = "caterpillar",
               values_to = "weight",
               values_drop_na = TRUE)

#make summary
cat_summary <- cat_long %>% 
            group_by(Experiment, soil_inocula, ID) %>%
  summarise(
    n = n(),
    mean = mean(weight, na.rm = TRUE),
    sd = sd(weight, na.rm = TRUE)) %>%
    mutate(se = sd/sqrt(n))


```

##Plot
```{r}

#plot boxplot Phytohormone-treated
phyto <- ggplot(data = subset(cat_summary, Experiment == "simulated"),
       mapping = aes(x = soil_inocula, y = mean), colour= soil_inocula)+
  geom_boxplot(mapping = aes(fill= soil_inocula), position = position_dodge(width = 0.8), lwd=0.8)+
  geom_jitter(data = subset(cat_summary, Experiment == "simulated"),
              mapping = (aes(soil_inocula, mean, fill = soil_inocula)),
              shape = 21, size = 1.8, color = "black",
              position = position_jitterdodge(jitter.height = .1, jitter.width = .2))+
  scale_fill_manual(values = color_pathway_induced)+
  ylim(0,35)+
  axis_looks+
  theme(legend.position = "none")+
  labs(y="M. brassicae weight (mg) 8dpi", x=" ")

#plot boxplot Herbivore-infested
insects <- ggplot(data = subset(cat_summary, Experiment == "real_insects"),
       mapping = aes(x = soil_inocula, y = mean), colour= soil_inocula)+
  geom_boxplot(mapping = aes(fill= soil_inocula), position = position_dodge(width = 0.8), lwd=0.8)+
  geom_jitter(data = subset(cat_summary, Experiment == "real_insects"),
              mapping = (aes(soil_inocula, mean, fill = soil_inocula)),
              shape = 21, size = 1.8, color = "black",
              position = position_jitterdodge(jitter.height = .1, jitter.width = .2))+
  scale_fill_manual(values = color_pathway_induced)+
  axis_looks+
  ylim(0,35)+
  theme(legend.position = "none")+
  labs(y="M. brassicae weight (mg) 8dpi", x=" ")

#combine plots
library(patchwork)
combined_plot <- phyto/insects

#check
combined_plot

```

##Stats
```{r}

#Phytohormone-treated
phyto <- subset(cat_summary, Experiment == "simulated")
phyto %>% group_by(Experiment,soil_inocula) %>% summarise(n=n()) #not-balanced

#lm
lm.p <- glm(mean ~ soil_inocula, 
            family=gaussian,
         data=phyto)

lm.p_table <- car::Anova(lm.p, type="II", test.statistic="LR") #no effect of soil treatment
simulateResiduals(lm.p, plot=T) #ok

#Herbivore-infested
insects <- subset(cat_summary, Experiment == "real_insects")
insects %>% group_by(Experiment,soil_inocula) %>% summarise(n=n()) #balanced


#glmer
lm.i <- glm(mean ~ soil_inocula, 
            family=gaussian,
         data=insects)

lm.i_table <- car::Anova(lm.i, type="II", test.statistic="LR") #no effect of soil treatment
simulateResiduals(lm.i, plot=T) #ok


```

#(d)Aphid population development
##Data
```{r}

#load data
aphids <- readRDS("./Data/plant_data/aphids.rds")

#change levels of Experiment & soil treatment
aphids <- aphids %>%
                mutate(Experiment = factor(Experiment,
                                           levels = c("simulated", "real_insects" )))%>%
                mutate(soil_inocula = factor(soil_inocula,
                                                      levels=c("control-conditioned",
                                                               "JA-conditioned", "SA-conditioned")))


#remove plants with wrong infestation
aphids <- aphids %>% 
          subset(ID != 145 & ID!= 351)%>% #removed 2 plants with wrong infestation 
                   droplevels()

#make it long
aphids_long <- aphids %>% 
               tidyr::pivot_longer(cols = c(7:8), 
               names_to = "time_point",
               values_to = "count",
               values_drop_na = TRUE)

#change to factor
aphids_long[c(1:2,7)] <- lapply(aphids_long[c(1:2,7)], as.factor)

#order time points
aphids_long$time_point <- factor(aphids_long$time_point,
                                 levels = c("dpi_8", "dpi_14")) 


```

##Plot
```{r}

#Phytohormone-treated
aphids_8dpi <- ggplot(data = subset(aphids_long, Experiment == "simulated" & time_point == "dpi_8"),
                 mapping = aes(x = soil_inocula, y = count), 
              colour= soil_inocula)+
  geom_boxplot(mapping = aes(fill= soil_inocula), position = position_dodge(width = 0.8), lwd=0.8)+
  geom_jitter(data = subset(aphids_long, Experiment == "simulated" & time_point == "dpi_8"),
              mapping = (aes(soil_inocula, count, fill = soil_inocula)),
              shape = 21, size = 1.8, alpha=0.9, color = "black",
              position = position_jitterdodge(jitter.height = .1, jitter.width = .2))+
  scale_fill_manual(values = color_pathway_induced)+
  axis_looks+
  ylim(0,80)+
  #facet_wrap(~time_point, scales="free_y")+
  theme(legend.position="none")

aphids_14dpi <- ggplot(data = subset(aphids_long, Experiment == "simulated" & time_point == "dpi_14"),
                 mapping = aes(x = soil_inocula, y = count), 
              colour= soil_inocula)+
  geom_boxplot(mapping = aes(fill= soil_inocula), position = position_dodge(width = 0.8), lwd=0.8)+
  geom_jitter(data = subset(aphids_long, Experiment == "simulated" & time_point == "dpi_14"),
              mapping = (aes(soil_inocula, count, fill = soil_inocula)),
              shape = 21, size = 1.8, alpha=0.9,color = "black",
              position = position_jitterdodge(jitter.height = .1, jitter.width = .2))+
  scale_fill_manual(values = color_pathway_induced)+
  axis_looks+
  ylim(0,600)+
  #facet_wrap(~time_point, scales="free_y")+
  theme(legend.position="none")


#Herbivore-infested
aphids_8dpi_i <- ggplot(data = subset(aphids_long, Experiment == "real_insects" & time_point == "dpi_8"),
                 mapping = aes(x = soil_inocula, y = count), 
              colour= soil_inocula)+
  geom_boxplot(mapping = aes(fill= soil_inocula), position = position_dodge(width = 0.8), lwd=0.8)+
  geom_jitter(data = subset(aphids_long, Experiment == "real_insects" & time_point == "dpi_8"),
              mapping = (aes(soil_inocula, count, fill = soil_inocula)),
              shape = 21, size = 1.8, alpha=0.9,color = "black",
              position = position_jitterdodge(jitter.height = .1, jitter.width = .2))+
  scale_fill_manual(values = color_pathway_induced)+
  axis_looks+
  ylim(0,80)+
  #facet_wrap(~time_point, scales="free_y")+
  theme(legend.position="none")

aphids_14dpi_i <- ggplot(data = subset(aphids_long, Experiment == "real_insects" & time_point == "dpi_14"),
                 mapping = aes(x = soil_inocula, y = count), 
              colour= soil_inocula)+
  geom_boxplot(mapping = aes(fill= soil_inocula), position = position_dodge(width = 0.8), lwd=0.8)+
  geom_jitter(data = subset(aphids_long, Experiment == "real_insects" & time_point == "dpi_14"),
              mapping = (aes(soil_inocula, count, fill = soil_inocula)),
              shape = 21, size = 1.8, alpha=0.9,color = "black",
              position = position_jitterdodge(jitter.height = .1, jitter.width = .2))+
  scale_fill_manual(values = color_pathway_induced)+
  axis_looks+
  ylim(0,600)+
  #facet_wrap(~time_point, scales="free_y")+
  theme(legend.position="none")

#combined plot
p.aphids <- (aphids_8dpi+aphids_14dpi)/(aphids_8dpi_i+aphids_14dpi_i)

#check
p.aphids

```


##Stats
```{r}

#full df
aphids_long

#Phytohormone-treated
phyto <- subset(aphids_long, Experiment == "simulated")
phyto %>% group_by(Experiment,soil_inocula) %>% summarise(n=n()) 

#run glm
glm.p <- glm(count ~ soil_inocula*time_point,
              family=Gamma(link="log"),
              data=phyto)

#check model assumptions
simulateResiduals(glm.p , plot=T) #ok

#get summary
car::Anova(glm.p , type="II", test.statistic="LR") #sig for timepoint and marginally sig for soil treatment, no interaction


#Herbivore-infested
insects <- subset(aphids_long, Experiment == "real_insects")
insects %>% group_by(Experiment,soil_inocula) %>% summarise(n=n())

#run glm
glm.i <- glm(count ~ soil_inocula*time_point,
              family=Gamma(link="log"),
              data=insects)

#check model assumptions
simulateResiduals(glm.i, plot=T) #ok

#get summary
car::Anova(glm.i , type="II", test.statistic="LR") #sig for timepoint, non-significant for soil treatment nor interaction. 

```

